Deciding Bitvector Termination with Program Synthesis

### 2014-06-25 papers/termination/paper.tex

Figure~\ref{fig:motivation.c} invalidates assumption (iii). 
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.

@@

Figure~\ref{fig:motivation.g} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, 
where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a disjunctively well-founded one, which requires an expensive binary 
reachability analysis.
However, by using our technique we can find a non-linear lexicographic ranking functions.
A lexicographic non-linear ranking functions consists of lexicographically ordered components
of non-linear functions. As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
loop transition leads to a decrease with respect to the lexicographic
ordering for this tuple. Therefore no function may increase unless a function of
a lower index decreases. Additionally, at every step, there must be at least one
function that decreases.

When proving program termination we are simultaneously solving two problems:
the search for a termination argument, and the search for a supporting invariant.
In our setting, a candidate termination argument is iteratively constructed. 
A candidate termination argument is composed out of a ranking function 
and a supporting invariant.

The program in Figure~\ref{fig:motivation.i} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
\begin{verbatim}
R(x, y, z) = (x < y, z)
\end{verbatim}

@@

\subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
We represent programs as transition systems with state space $X$, transition relation $T \subseteq X \times X$
and the set of variables $$

\begin{definition}[Unconditional termination]
A transition system with state space $X$ and transition relation $T \subseteq X \times X$
is said to be \emph{unconditionally terminating} if there is no infinite sequence
$\vec{x_1}, \vec{x_2}, \ldots$ with $\forall i . T(\vec{x_i}, \vec{x_{i+1}})$.  
\end{definition}

@@

\begin{definition}[Ranking function]
Given a transition system with state space $X$ and transition relation $T \subseteq X \times X$,
a ranking function is an injective function $R: X \to Y$ where
$Y$ is well-founded and $R$ is monotonically decreasing with respect to $T$.  That is
to say:
$$\forall \vec{x}, \vec{x'} . T(\vec{x}, \vec{x'}) \Rightarrow R(\vec{x}) < R(\vec{x'})$$
\end{definition}

\begin{definition}[Linear ranking function]
A linear ranking function $R: X \to Y$ 
with $\dim(X) = n$ and $\dim(Y) = m$ is of the form:
is one that can be expressed as an $n \times m$ matrix $M$:
$$f(\vec{x}) = M\vec{x}$$
\end{definition}

\begin{definition}[Lexicographic ranking function]
If $Y = Z^m$, we say that a ranking function $R: X \to Y$ is lexicographic
and maps each state in $X$ to a tuple of values such that the loop transition leads to a decrease with
respect to the lexicographic ordering for this tuple.
The total order imposed on $Y$ is the lexicographic ordering
induced on tuples of $Z$'s. 
\end{definition}

For lexicographic ranking functions, no function may increase unless a function of a lower index decreases.
Additionally, at every step, there must be at least one function that decreases.
We note that some termination arguments require lexicographic ranking functions, or equivalently, ranking functions
whose co-domain is the ordinals, rather than just $\mathbb{N}$.

In the rest of the paper, we consider any ranking function $R: X \to Z^m$ to be lexicographic, and the situation when $m=1$ to be a special case.

A special case of a ranking function is a \emph{linear ranking function}.  This
class of functions is exactly what you'd expect: a linear function that satisfies
the criteria for a ranking function.  We recall that a linear function $f: X \to Y$,
with $\dim(X) = n$ and $\dim(Y) = m$
is one that can be expressed as an $n \times m$ matrix $M$:
$$f(\vec{x}) = M\vec{x}$$

@@

@@

\begin{figure*}
\centering
\begin{tabular}{|l|r|c|c|c|c|r|r|r|}
\hline
    & LOC & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional? & Float? & Dimension & \shortstack{Ranking \\ program size} & Time (s)\\
    \hline
    \hline
 P1 & 10 & \xmark & \xmark & \xmark & \xmark & 3 & 1 & 0.01 \\
 P2 & 10 & \xmark & \xmark & \xmark & \xmark & 3 & 1 & 0.01 \\
 P3 & 10 & \xmark & \xmark & \xmark & \xmark & 3 & 1 & 0.01 \\
 \hline
\end{tabular}

>>>>

Figure~\ref{fig:motivation.c} invalidates assumption (iii). 
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic as the loop condition stays always true for any initial \texttt{x} at least 1.

@@

Up until this point, we looked at examples that are not soundly treated by existing techniques as they 
don't fit in the range of programs covered by these techniques.
Next, we look at some programs that are handled by existing termination tools via dedicated analyses. We show that our method 
uniformly handles them, without the need for any special treatment.

Figure~\ref{fig:motivation.g} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, 
where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a Ramsey-based termination argument,
which requires an expensive binary reachability analysis.
With our technique we can find a non-linear lexicographic ranking function $R(x) = ...$.
A lexicographic non-linear ranking functions consists of lexicographically ordered components
of non-linear functions. 
%As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
%loop transition leads to a decrease with respect to the lexicographic
%ordering for this tuple. Therefore no function may increase unless a function of
%a lower index decreases. Additionally, at every step, there must be at least one
%function that decreases.

The loop in Figure~\ref{fig:motivation.i} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multiphase ranking function computed from a multiphase ranking template.
A multiphase ranking template is targeted at programs that go through a
finite number of phases in their execution. Each phase is ranked with
an affine-linear function and the phase is considered to be completed once this
function becomes non-positive.
However, in our setting this type of programs do not need a special treatment, as we can find a non-linear ranking function:
\begin{verbatim}
R(x, y, z) = (x < y, z)
\end{verbatim}

The loop in Figure~\ref{fig:motivation.h} illustrates conditional termination.
Basically, when proving program termination we are simultaneously solving two problems:
the search for a termination argument, and the search for a supporting invariant.
We iteratively construct a termination argument as a pair of a ranking function and supporting invariant through a single refinement loop.
For this loop, we find the ranking function $R(x) = x$ together with the supporting invariant $y=1$.

@@

\subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
We represent programs as transition systems with state space $X$ and transition relation $T \subseteq X \times X$.

\begin{definition}[Unconditional termination]
A transition system is said to be \emph{unconditionally terminating} if there is no infinite sequence
$x_1, x_2, \ldots$ with $\forall i . T(x_i, x_{i+1})$.  
\end{definition}

@@

\begin{definition}[Ranking function]
An injective function $R: X \to Y$ is a ranking function if $Y$ is well-founded and 
$R$ is monotonically decreasing with respect to $T$.  That is
to say:
$$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$
\end{definition}

\begin{definition}[Linear ranking function]
A linear ranking function $R: X \to Y$ 
with $\dim(X) = n$ and $\dim(Y) = m$ is of the form: $$f(\vec{x}) = M\vec{x}$$ where
$M$ is an $n \times m$ matrix.
\end{definition}

@@

\begin{definition}[Lexicographic ranking function]
For $Y = Z^m$, we say that a ranking function $R: X \to Y$ is lexicographic
if it maps each state in $X$ to a tuple of values such that the loop transition leads to a decrease with
respect to the lexicographic ordering for this tuple.
The total order imposed on $Y$ is the lexicographic ordering
induced on tuples of $Z$'s. 
\end{definition}

For lexicographic ranking functions, no function may increase unless a function of a lower index decreases.
Additionally, at every step, there must be at least one function that decreases.
We note that some termination arguments require lexicographic ranking functions, or equivalently, ranking functions
whose co-domain is the ordinals, rather than just $\mathbb{N}$.
%In the rest of the paper, we consider any ranking function $R: X \to Z^m$ to be lexicographic, and the situation when $m=1$ to be a special case.

\begin{definition}{Conditional termination}
\end{definition}

\begin{definition}{Supportive invariant}
\end{definition}

@@

\begin{figure*}
\centering
\begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|}
\hline
    & LOC & Terminates? & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional? & Float? & Dimension & \shortstack{Ranking \\ program size} & Res & Time (s)\\
    \hline
    \hline
\input{table}
    \hline
\end{tabular}
